%Version 3.1 December 2024
% See section 11 of the User Manual for version history
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%                                                                 %%
%% Please do not use \input{...} to include other tex files.       %%
%% Submit your LaTeX manuscript as one .tex document.              %%
%%                                                                 %%
%% All additional figures and files should be attached             %%
%% separately and not embedded in the \TeX\ document itself.       %%
%%                                                                 %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass[pdflatex,sn-mathphys-num]{sn-jnl}% Math and Physical Sciences Numbered Reference Style

%%%% Standard Packages
\usepackage{graphicx}%
\usepackage{multirow}%
\usepackage{amsmath,amssymb,amsfonts}%
\usepackage{amsthm}%
\usepackage{mathrsfs}%
\usepackage[title]{appendix}%
\usepackage{xcolor}%
\usepackage{textcomp}%
\usepackage{manyfoot}%
\usepackage{booktabs}%
\usepackage{algorithm}%
\usepackage{algorithmicx}%
\usepackage{algpseudocode}%
\usepackage{listings}%
%%%%

\theoremstyle{thmstyleone}%
\newtheorem{theorem}{Theorem}%
\newtheorem{proposition}[theorem]{Proposition}%

\theoremstyle{thmstyletwo}%
\newtheorem{example}{Example}%
\newtheorem{remark}{Remark}%

\theoremstyle{thmstylethree}%
\newtheorem{definition}{Definition}%

\raggedbottom

\begin{document}

\title[GPU Acceleration of EKF Fusion Positioning for Cased-Hole Robots]{Taming the Latency-Bound Beast: GPU Acceleration of EKF Fusion Positioning for Cased-Hole Robots via Kernel Fusion, Mixed Precision, and Multi-Trajectory Parallelism}

\author*[1]{\fnm{forename} \sur{surname}}\email{author@example.com}

\affil*[1]{\orgdiv{Department}, \orgname{University}, \orgaddress{\city{City}, \state{State}, \country{China}}}

\abstract{Extended Kalman filter (EKF) fusion positioning for cased-hole robots presents a worst-case workload for GPU acceleration: a strictly sequential 15-state recursion over 166,667 time steps, with per-step matrices no larger than $15\times15$ and over 500 micro-kernel launches per step in naive implementations. We show that naive GPU execution is 4$\times$ slower than single-core CPU, with SM throughput below 0.3\%---a textbook latency-bound problem where compute units idle waiting on dependencies rather than feeding on arithmetic. We address this through three orthogonal optimizations: (1) \textbf{kernel fusion} via a single Triton mega-kernel that collapses 531 CUDA kernel launches into one, achieving 8.8$\times$ speedup over CPU in fp64; (2) \textbf{mixed-precision execution} that leverages fp32 to deliver a further 14.3$\times$ over the fused fp64 kernel (133$\times$ vs.~CPU) while maintaining sub-millimeter positioning accuracy; (3) \textbf{multi-trajectory batch parallelism} that maps independent trajectory scans to a CUDA grid, achieving perfect linear scaling up to the SM count (78$\times$ at 78 trajectories) and 18,210$\times$ aggregate throughput at batch size 256 (34 million steps/s). We also present negative results showing that Tensor Cores, TMA, warp specialization, and conventional tile/layout auto-tuning offer zero or negative benefit for this workload class, providing guidance for practitioners facing similar latency-bound, micro-matrix recursive workloads.}

\keywords{Cased-hole robot, Fusion positioning, Extended Kalman filter, GPU kernel fusion, Mixed precision, Multi-trajectory parallelism, Triton}

\maketitle

%% ============================================================
%% Section 1: Introduction
%% ============================================================
\section{Introduction}\label{sec1}

\subsection{Motivation}\label{subsec:motivation}

In horizontal and extended-reach well logging, cased-hole traction robots perform autonomous traversal, instrument delivery, and station-keeping in environments devoid of GPS, with unreliable geomagnetic references due to metallic casing interference. Multi-sensor fusion combining strapdown inertial navigation (SINS) and dual wheel odometers via an extended Kalman filter (EKF) has emerged as the primary positioning approach~\cite{bib1}.

The simulation system for validating such fusion algorithms must generate long-distance trajectories with random obstacle perturbations, simulate IMU and odometer sensor data, execute SINS mechanization, and run a 15-state EKF over hundreds of thousands of sampling points. As simulation distance, sampling frequency, batch experiment count, and parameter search scale increase, computational performance becomes a critical bottleneck.

We study a concrete instance: a 500\,m pipeline trajectory at 100\,Hz with 166,667 sampling points, processed by a 15-state SINS/EKF fusion filter. On the surface, this appears trivially small---each EKF step involves only $15\times15$ matrix operations and a handful of transcendental functions, totaling approximately 25,000 FLOPs per step. Yet, as we show, this workload is maximally hostile to GPU acceleration: the sequential recursion prevents temporal parallelism, the micro-matrices cannot saturate GPU compute units, and the fragmentation of operations into hundreds of micro-kernels per step creates launch and synchronization overhead that dwarfs the actual computation.

\subsection{The Latency-Bound Challenge}\label{subsec:challenge}

Contemporary GPU optimization literature focuses overwhelmingly on throughput-bound workloads: large matrix multiplications, attention mechanisms, and convolutional kernels where the primary challenge is feeding enough arithmetic to saturate compute units~\cite{bib2}. Our workload inverts this pattern. With a single trajectory, we measure:

\begin{itemize}
\item \textbf{SM throughput of 0.27\%}: Only 4 out of 64 warps per streaming multiprocessor are active, and the single CUDA block occupies just 1 of 78 SMs on an NVIDIA H100-class GPU.
\item \textbf{531 CUDA kernel launches per EKF step}: A naive PyTorch implementation fragments each 25k-FLOP step into 531 micro-kernels (element-wise multiply, subtract, concatenate, negate), with only $\sim$20\% of GPU time spent in actual linear algebra (cuBLAS GEMM).
\item \textbf{Naive GPU is 4$\times$ slower than CPU}: At 263 steps/s on GPU versus 1,866 steps/s on a single CPU core, the GPU implementation loses catastrophically to the CPU---a rare and instructive failure mode.
\end{itemize}

This is a \textit{latency-bound} problem: the GPU's compute units spend most of their time idle, waiting on kernel launch dispatch, inter-kernel synchronization, and the sequential dependency chain of the Kalman recursion. The challenge is not ``how to compute more'' but ``how to eliminate the overhead that prevents us from computing at all.''

\subsection{Contributions}\label{subsec:contributions}

We address this challenge through three orthogonal optimizations that, when multiplied, transform a GPU-hostile workload into one that exploits GPU hardware effectively:

\begin{itemize}
\item \textbf{Operator-level: Single-Kernel Fusion (v1--v3).} We progressively eliminate fragmentation: first removing Python-scalar synchronization and graph breaks (v1, 1.12$\times$), then capturing the per-step graph with CUDA Graphs to eliminate CPU launch overhead (v2, 1.41$\times$), and finally writing the entire sequential scan as a single Triton kernel with state resident in registers across all 166,667 steps (v3, 8.8$\times$ over CPU in fp64). This reduces 531 kernel launches per step to \textbf{1 launch for the entire trajectory}.

\item \textbf{Algorithm-level: Mixed-Precision Execution (v4).} By parameterizing the mega-kernel's data type as a compile-time constant and switching from fp64 to fp32, we achieve an additional 14.3$\times$ speedup over the fused fp64 kernel (133$\times$ vs.~CPU baseline), while maintaining sub-millimeter positioning accuracy ($\Delta X \approx 0.11$\,mm). We further show that Tensor Core acceleration (tf32) is \textit{counterproductive} for this workload: it is slower than fp32 and introduces centimeter-level errors, due to the overhead of padding $15\times15$ matrices to TC-compatible tile sizes.

\item \textbf{System-level: Multi-Trajectory Batch Parallelism (v5).} Since a single trajectory's sequential recursion cannot be parallelized along the time dimension, we exploit \textit{inter}-trajectory parallelism by mapping independent trajectories to distinct CUDA blocks. This achieves \textbf{perfect linear scaling} up to the SM count (78$\times$ at 78 trajectories), with aggregate throughput reaching 34 million steps/s at batch size 256 (18,210$\times$ over CPU baseline). SM utilization rises from 0.27\% to 19.91\%.
\end{itemize}

We also contribute a systematic \textbf{negative-results section} demonstrating that several SOTA GPU optimization techniques---Tensor Cores (via tf32), TMA (Tensor Memory Accelerator), warp specialization, and conventional tile/layout auto-tuning---provide zero or negative benefit for this workload class. These findings establish that the optimization space for latency-bound micro-matrix recursive workloads is structurally different from that of throughput-bound large-matrix workloads.

%% ============================================================
%% Section 2: Background
%% ============================================================
\section{Background}\label{sec2}

\subsection{Cased-Hole Robot Positioning System}\label{subsec:system}

\subsubsection{Hardware Configuration and Environment}

The cased-hole robot operates within a sealed metallic pipeline (inner diameter 130\,mm, robot outer diameter 80\,mm, single-side radial clearance 25\,mm). The sensor suite comprises a three-axis gyroscope (noise $\sigma = 0.005$\,rad/s), a three-axis accelerometer (noise $\sigma = 0.01$\,m/s$^2$), and dual-path wheel odometers (noise $\sigma = 0.02$\,m/s). Four coordinate frames are defined: the IMU frame (coincident with the robot body frame), the odometer frame, and the navigation frame (local-level, x-forward, y-right, z-up).

\subsubsection{Fusion Algorithm}

The positioning scheme employs SINS as the primary navigation system and dual odometers as auxiliary sensors within an EKF framework~\cite{bib3}. The 15-dimensional error state vector comprises:

\begin{equation}
\mathbf{x} = [\delta\mathbf{p}^T, \delta\mathbf{v}^T, \boldsymbol{\phi}^T, \boldsymbol{\varepsilon}^T, \nabla^T]^T \in \mathbb{R}^{15}
\label{eq:state}
\end{equation}

where $\delta\mathbf{p}$, $\delta\mathbf{v}$, $\boldsymbol{\phi}$ are 3D position, velocity, and misalignment errors, and $\boldsymbol{\varepsilon}$, $\nabla$ are 3D gyroscope and accelerometer biases.

Each time step executes 7 sequential stages: (1) quaternion attitude update via gyroscope integration, (2) direction cosine matrix (DCM) conversion from updated quaternion, (3) SINS velocity and position integration, (4) dual-odometer displacement averaging, (5) anomaly detection via SINS-odometer displacement comparison, (6) EKF prediction with state transition matrix $F$ and covariance propagation $P^- = FPF^T + Q$, followed by Kalman gain computation and state update, and (7) error compensation on position, velocity, and attitude. The observation model uses a fixed 3-dimensional measurement (displacement difference plus Y/Z velocity constraints), with the odometer channel suppressed via $R_{\text{odo}} \rightarrow 10^{12}$ under anomaly---mathematically equivalent to dropping the observation row but structurally uniform for GPU execution.

\subsection{Simulation Dataset}\label{subsec:dataset}

The simulation constructs a 500\,m trajectory at 100\,Hz with 0.3\,m/s constant forward velocity, yielding $N = 166{,}667$ sampling points. Obstacles (scaling, corrosion, couplings) are randomly generated every 20--40\,m starting from 50\,m, with heights of 1--2\,mm and Gaussian-profile influence range of $\pm 0.5$\,m. The output dataset contains 15 structured fields: timestamp, 3D position truth, 3D velocity truth, 3-axis angular velocity, 3-axis acceleration, and dual odometer velocities.

The CPU fp64 baseline achieves the following positioning accuracy, which serves as the golden standard for all subsequent comparisons:

\begin{table}[h]
\caption{Baseline positioning accuracy (CPU fp64, $N=166{,}667$)}\label{tab:baseline}
\begin{tabular}{@{}lccc@{}}
\toprule
Metric & X-axis & Y-axis & Z-axis \\
\midrule
RMSE (mm) & 1017.84 & 16.83 & 8.14 \\
Relative accuracy $\rho$ & \multicolumn{3}{c}{0.1755\%} \\
Closure error (mm) & \multicolumn{3}{c}{877.70} \\
\botrule
\end{tabular}
\end{table}

\subsection{Why This Workload is GPU-Hostile}\label{subsec:gpu_hostile}

We identify four compounding factors that make this workload maximally hostile to GPU acceleration:

\begin{enumerate}
\item \textbf{Sequential dependency.} The Kalman recursion imposes strict step-to-step dependence: state $(pos, vel, q, P)$ at step $k$ is a function of step $k-1$. The time dimension admits zero parallelism, ruling out batched or pipelined execution along the temporal axis.

\item \textbf{Micro-matrix arithmetic.} All involved matrices are $\leq 15\times15$ (padded to $16\times16$ for Triton's \texttt{tl.dot}). A single matrix multiply of this size engages at most a fraction of one warp, yielding compute utilization below 1\%. The per-step true FLOP count is approximately 25,092 (eight $\leq 15\times15$ GEMMs, one $3\times3$ analytic inverse, and scalar transcendental functions), yet the naive implementation requires 531 kernel launches.

\item \textbf{Operator fragmentation.} Profiling the CPU eager implementation reveals 1,026 PyTorch operator calls per step (2.05 million total over 2,000 steps). True linear algebra (matrix inverse, solve, multiply) accounts for only 15--20\% of execution time. The dominant costs are fragmented scalar operations: \texttt{aten::select} (14.2\%), \texttt{aten::mul} (11.8\%), and \texttt{aten::item}/\texttt{\_local\_scalar\_dense} ($\sim$5\% each), all stemming from per-element tensor construction of the state transition matrix $F$, observation matrix $H$, and quaternion/DCM/skew-symmetric helper matrices.

\item \textbf{Latency-bound execution.} On GPU, the 531 micro-kernels per step each incur $\sim 0.7\,\mu$s of inter-kernel transition latency. The cumulative effect is a per-step GPU time of $\sim 817\,\mu$s dominated by dispatch and synchronization overhead, with actual computation occupying a small fraction. The NVIDIA Nsight Compute measurement confirms: \texttt{sm\_\_throughput} = 0.27\%, \texttt{sm\_\_warps\_active} = 6.25\% (4/64 warps), and only 1 of 78 SMs occupied.
\end{enumerate}

This characterization is essential because it determines the viable optimization space: traditional throughput-oriented techniques (larger tiles, higher occupancy, Tensor Cores) address a different problem. Our problem demands latency reduction through kernel fusion, precision reduction to match hardware throughput ratios, and multi-trajectory parallelism to fill idle SMs.

\subsection{Experimental Platform}\label{subsec:platform}

All experiments are conducted on an NVIDIA Hopper architecture GPU (sm\_90) with 78 streaming multiprocessors and 150\,GB HBM, running PyTorch 2.11.0 with CUDA 12.9 and Triton 3.6.0. The CPU baseline uses a single core of the host processor.

%% ============================================================
%% Section 3: Method
%% ============================================================
\section{Method}\label{sec3}

\subsection{Design Overview}\label{subsec:overview}

Our optimization proceeds along three orthogonal axes, each attacking a distinct aspect of the latency-bound micro-matrix recursive workload:

\begin{itemize}
\item \textbf{Axis 1---Operator Fusion (v1$\rightarrow$v2$\rightarrow$v3):} Eliminate kernel launch and synchronization overhead by progressively fusing fragmented operations, culminating in a single Triton kernel that executes the entire 166,667-step sequential scan.

\item \textbf{Axis 2---Mixed Precision (v4):} Exploit the large fp64-to-fp32 throughput ratio on modern GPUs by switching the fused kernel to fp32, with careful validation that positioning accuracy remains sub-millimeter.

\item \textbf{Axis 3---Multi-Trajectory Batch (v5):} Since single-trajectory latency has a hard lower bound (the sequential recursion), achieve throughput scaling by mapping independent trajectories to distinct CUDA blocks, filling idle SMs.
\end{itemize}

These three axes are \textit{multiplicative}: the final aggregate throughput is approximately the product of the speedups from fusion, mixed precision, and batch parallelism. Table~\ref{tab:versions} summarizes the version evolution.

\begin{table}[h]
\caption{Version evolution: throughput, speedup, and correctness (full $N=166{,}667$)}\label{tab:versions}
\begin{tabular}{@{}llrrcc@{}}
\toprule
Version & Method & Device/Precision & Throughput (steps/s) & vs.~v0 CPU & Correctness \\
\midrule
v0 & Naive eager & CPU fp64 & 1,866 & 1.00$\times$ & Golden \\
v0 & Naive eager & GPU fp64 & 263* & 0.14$\times$ & (slower than CPU) \\
v1 & Fragment elimination & CPU fp64 & 2,098 & 1.12$\times$ & Bit-exact \\
v2 & CUDA Graph & GPU fp64 & 2,640 & 1.41$\times$ & Bit-exact \\
v3 & Triton mega-kernel & GPU fp64 & 16,491 & 8.8$\times$ & Sub-mm \\
\textbf{v4} & \textbf{+ fp32} & \textbf{GPU fp32} & \textbf{242,737} & \textbf{130$\times$} & \textbf{Sub-mm} \\
v4 & + tf32 & GPU tf32 & 209,524 & 112$\times$ & Sub-cm \\
\textbf{v5} & \textbf{+ Batch$\times$256} & \textbf{GPU fp32} & \textbf{34.0M} & \textbf{18,210$\times$} & \textbf{Bit-exact} \\
\botrule
\end{tabular}

\smallskip
\footnotesize{*v0 GPU extrapolated from 20,001-step partial run. v5 throughput is aggregate across 256 independent trajectories; single-trajectory latency is unchanged from v4.}
\end{table}

\subsection{Axis 1: Operator Fusion---From 531 Kernels to 1}\label{subsec:fusion}

\subsubsection{v1: Fragment Elimination and Graph-Break Removal}

The first stage targets the root cause identified by profiling: 1,026 operator calls per step, dominated by Python-scalar synchronization (\texttt{.item()}) and per-element tensor construction. We rewrite the EKF step function to:

\begin{itemize}
\item Construct all matrices ($\theta_{\text{mat}}$, $C_n^b$, $H$, $F$, skew-symmetric blocks) using \texttt{torch.stack} and \texttt{torch.cat} on GPU tensors, eliminating all Python-scalar intermediates and \texttt{.item()} calls.
\item Unify the observation model to a fixed 3-dimensional form: under odometer anomaly, set $R_{\text{odo}} \rightarrow 10^{12}$ to suppress the odometer channel (mathematically equivalent to deleting the observation row via $K \approx 0$).
\item Replace conditional branches with \texttt{torch.where} and \texttt{clamp} operations.
\end{itemize}

PyTorch's \texttt{torch.\_dynamo.explain} confirms the result: \textbf{graph\_count = 1, graph\_break\_count = 0, op\_count = 241} (vs.~1,026 in v0, a 76\% reduction). This yields a modest 1.12$\times$ CPU speedup. Critically, we discover that \texttt{torch.compile(default)} is \textit{slower} than eager execution for this workload: the single-step graph is too small for Inductor's code generation and guard-checking overhead to be amortized.

\subsubsection{v2: CUDA Graph Capture/Replay}

To eliminate the remaining CPU-side kernel launch scheduling overhead, we capture the per-step computation graph as a CUDA Graph and replay it for all $N$ steps. Key implementation challenges:

\begin{itemize}
\item \texttt{torch.linalg.inv} (which internally calls cuSOLVER) triggers \texttt{cudaErrorStreamCaptureInvalidated} during graph capture due to internal host synchronization. We replace it with an analytic $3\times3$ adjugate-based inverse: $\text{inv}(S) = \text{adj}(S) / \det(S)$, which is purely element-wise.
\item GPU-scalar tensor indexing triggers implicit host synchronization during capture. We replace \texttt{gyro[idx]} with \texttt{torch.index\_select} on a 1-element index tensor.
\end{itemize}

The CUDA Graph replay achieves 2,640 steps/s (1.41$\times$ over v0 CPU). However, \textbf{unrolling} the graph (capturing 1/5/10/20/50 steps per graph) yields no additional speedup, confirming that the bottleneck has shifted from CPU launch overhead to GPU execution---the 531 micro-kernels within each step still execute sequentially with inter-kernel transition delays.

\subsubsection{v3: Triton Mega-Kernel---The Core Contribution}

The key insight from v2 is that \textit{531 dependent micro-kernels per step} is the true bottleneck. Each kernel incurs $\sim 0.7\,\mu$s of dispatch and synchronization overhead, and these delays accumulate linearly over the step's dependency chain. The solution is to eliminate all intra-step kernel boundaries: write the \textbf{entire sequential scan as a single Triton kernel}.

Our mega-kernel design:

\begin{itemize}
\item \textbf{Single launch, single block.} The kernel is launched once with \texttt{grid=(1,)} and executes all $N = 166{,}667$ steps in a \texttt{for} loop within the kernel body. This eliminates $N \times 531$ kernel launches.
\item \textbf{State resident in registers.} All persistent state---position (3), velocity (3), quaternion (4), covariance matrix $P$ ($16\times16$ padded), and the 15-element error state---is held in Triton tensors that the compiler maps to registers. There are zero global memory round-trips for state between steps.
\item \textbf{Padded-16 matrix operations.} All matrices are padded to $16\times16$ to use \texttt{tl.dot} with fp64 precision. The $3\times3$ matrix inverse uses analytic adjugate/determinant computation.
\item \textbf{Transcendental functions via libdevice.} \texttt{atan2}, \texttt{sin}, \texttt{cos}, \texttt{sqrt}, and \texttt{abs} use NVIDIA's libdevice implementations, which are directly callable from Triton.
\item \textbf{Correct $F$ matrix construction.} A critical implementation detail: $F[6{:}9,6{:}9] = -\text{skew}(\omega) \cdot dt$ \textit{overwrites} the diagonal block of the padded identity matrix, requiring explicit subtraction of the identity contribution before the skew-symmetric assignment. This subtle difference from additive accumulation caused an initial correctness bug.
\end{itemize}

\textbf{Performance:} 16,491 steps/s, representing \textbf{8.8$\times$} over CPU fp64 and \textbf{5.5$\times$} over the CUDA Graph implementation. The entire trajectory processes in 10.1 seconds.

\textbf{Correctness:} Verified via per-step position comparison against the CPU golden standard over 5,000 steps. Maximum absolute deviation: X = $3.8 \times 10^{-7}$\,m, Y = $1.6 \times 10^{-8}$\,m, Z = $9.0 \times 10^{-6}$\,m ($\leq 9\,\mu$m). These deviations are entirely attributable to floating-point accumulation order differences in fp64, three orders of magnitude below the RMSE (mm-scale).

\subsubsection{Why Not cuBLAS/cuSOLVER?}

A natural question is why we do not simply use NVIDIA's optimized libraries. We measured single-step reference latencies: cuBLAS GEMM ($F \cdot P \cdot F^T$, $15\times15$) requires 17.5\,$\mu$s, and cuSOLVER inverse ($3\times3$) requires 134\,$\mu$s. Remarkably, \textbf{cuSOLVER's single $3\times3$ inverse (134\,$\mu$s) is slower than our entire v3 mega-kernel step ($\sim$49\,$\mu$s)}. The reason is that these libraries are designed and optimized for large-matrix throughput, where fixed launch and metadata overhead amortizes over substantial computation. At $15\times15$ scale, the fixed cost dominates, rendering general-purpose libraries counterproductive.

\subsection{Axis 2: Mixed-Precision Execution}\label{subsec:mixed_precision}

\subsubsection{Method}

We parameterize the v3 mega-kernel with two compile-time constants: the data type \texttt{DT} and the \texttt{tl.dot} input precision \texttt{IP}. A single kernel source compiles to three precision variants:

\begin{itemize}
\item \textbf{fp64 (ieee):} Baseline fused kernel, identical to v3.
\item \textbf{fp32 (ieee):} Standard single-precision floating point.
\item \textbf{tf32 (Tensor Core):} \texttt{tl.dot} with \texttt{input\_precision=tf32}, using Tensor Core acceleration.
\end{itemize}

Truth data and final RMSE computation always use fp64 for fair accuracy comparison.

\subsubsection{Results}

\begin{table}[h]
\caption{Mixed-precision throughput and accuracy ($N=166{,}667$)}\label{tab:precision}
\begin{tabular}{@{}lrrrrc@{}}
\toprule
Precision & Throughput (steps/s) & vs.~fp64 kernel & vs.~v0 CPU & $\Delta$X (mm) & Accuracy \\
\midrule
fp64 & 17,090 & 1.0$\times$ & 9.3$\times$ & --- & Baseline \\
\textbf{fp32} & \textbf{243,587} & \textbf{14.3$\times$} & \textbf{133$\times$} & \textbf{0.11} & \textbf{Sub-mm} \\
tf32 & 209,472 & 12.3$\times$ & 114$\times$ & 4.71 & Sub-cm \\
\botrule
\end{tabular}
\end{table}

The fp32 result is a \textbf{14.3$\times$ speedup over the already-fused fp64 kernel}---far exceeding the expected 1.5--2$\times$ from the fp64:fp32 ALU throughput ratio alone. We attribute this super-linear gain to three compounding effects on the latency-bound critical path:

\begin{enumerate}
\item \textbf{Reduced register pressure}: fp64 values occupy two 32-bit registers each. Halving the register footprint for all persistent state allows the compiler to maintain more values in fast register file, reducing spill/fill overhead.
\item \textbf{Faster transcendentals}: libdevice \texttt{sin}, \texttt{cos}, \texttt{sqrt}, and \texttt{atan2} have significantly lower latency in fp32 than fp64 on Hopper architectures, and these functions lie on the critical path of quaternion update and DCM construction.
\item \textbf{Higher ALU throughput}: Hopper's fp32 ALU throughput is substantially higher than fp64, and in a latency-bound chain where every instruction contributes to the critical path, higher throughput directly translates to lower latency.
\end{enumerate}

\textbf{Accuracy:} fp32 introduces a maximum position deviation of $\Delta X \approx 0.11$\,mm, well within the millimeter-level positioning requirement. The X-axis RMSE of $\sim$1\,m is dominated by systematic integration drift, not numerical precision.

\textbf{Negative result---tf32/Tensor Cores:} The tf32 variant is \textit{slower} than fp32 (209k vs.~243k steps/s) and \textit{less accurate} ($\Delta X \approx 4.7$\,mm). At $15\times15$ (padded to $16\times16$), Tensor Core tiles waste significant computation on padding elements while introducing mantissa truncation errors. This demonstrates that Tensor Cores are structurally mismatched for micro-matrix workloads.

\subsection{Axis 3: Multi-Trajectory Batch Parallelism}\label{subsec:batch}

\subsubsection{Method}

Axes 1 and 2 minimize \textit{per-trajectory latency}, but the sequential recursion sets a hard lower bound on single-trajectory throughput. However, many practical scenarios involve \textbf{multiple independent trajectories}: Monte Carlo simulations with different random seeds, multi-sensor configurations, or multi-target tracking. These trajectories share identical computation structure but different input data, creating an opportunity for inter-trajectory parallelism.

We extend the v4 kernel by mapping the \textbf{trajectory dimension to the CUDA grid}: launching with \texttt{grid=(B,)} where each of the $B$ CUDA blocks independently processes one complete trajectory scan. The kernel body is unchanged from v4; only pointer offsets (\texttt{pid $\times$ batch\_stride}) differentiate trajectories. This design ensures zero inter-block communication and perfect isolation.

\subsubsection{Scaling Results}

\begin{table}[h]
\caption{Multi-trajectory batch scaling ($N=166{,}667$, fp32)}\label{tab:batch}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Batch $B$ & Throughput (steps/s) & Trajectories/s & Speedup vs.~$B{=}1$ & Notes \\
\midrule
1  & 229k       & 1.4   & 1.0$\times$   & Single trajectory (v4 fp32) \\
16 & 3.67M      & 22.0  & 16.0$\times$  & Perfect linear \\
32 & 7.35M      & 44.1  & 32.0$\times$  & Perfect linear \\
64 & 14.66M     & 88.0  & 64.0$\times$  & Perfect linear \\
78 & 17.86M     & 107.2 & 77.9$\times$  & $=$ SM count, 1 block/SM \\
128 & 24.07M    & 144.4 & 105$\times$    & 2nd wave (occupancy) \\
256 & \textbf{34.08M} & \textbf{204.5} & \textbf{148.7$\times$} & Near register limit \\
\botrule
\end{tabular}
\end{table}

\textbf{Perfect linear scaling at $B \leq 78$}: With one CUDA block per SM, speedup equals batch size exactly. This confirms that the trajectory-dimension mapping is the correct approach for saturating idle SMs.

\textbf{SM utilization evidence}: Nsight Compute measurements show \texttt{sm\_\_throughput} rising from 0.27\% (single block) to 19.91\% at $B=78$, a $\sim$74$\times$ increase consistent with 78 blocks occupying all 78 SMs. With 110 registers per thread, the occupancy limit is 4 blocks per SM, yielding a theoretical saturation point of $\sim$312 trajectories---consistent with the continued near-linear scaling observed at $B=256$.

\textbf{Inter-block correctness}: All trajectories produce bit-identical results (\texttt{traj\_spread} = 0), confirming zero cross-block interference.

It is important to note that v5's 34M steps/s is \textbf{aggregate throughput} across 256 trajectories, not single-trajectory acceleration. Single-trajectory latency remains bounded by the sequential recursion; Axes 1 and 2 have already extracted all available latency reduction. Axes 1+2 (latency) and Axis 3 (throughput) are orthogonal dimensions.

\subsection{Auto-Tuning and Advanced GPU Techniques: A Systematic Negative-Result Study}\label{subsec:negative}

A significant portion of our investigation was devoted to techniques that proved \textit{ineffective} for this workload. We document these negative results as guidance for practitioners and to establish the structural boundaries of the optimization space.

\subsubsection{Conventional Auto-Tuning}

We analyzed the input data characteristics: $dt$ is strictly uniform ($\sigma/\mu < 10^{-6}$), the batch size is 1 (single trajectory), and all tensor shapes remain static throughout the scan. The optimization dimensions traditionally explored by auto-tuning frameworks are:

\begin{itemize}
\item \textbf{Tile size}: Search space $\approx 0$. All matrices are $\leq 15\times15$; a single warp already covers the entire tile. Larger tiles would only add padding waste.
\item \textbf{Memory layout}: Search space $\approx 0$. Persistent state resides in registers; there is no tiled global/shared memory traffic to optimize.
\item \textbf{Kernel fusion strategy}: \textbf{Search space maximal}---the $531 \rightarrow 1$ kernel fusion (v3) is the only high-leverage direction, and it has been fully realized.
\item \textbf{Precision selection}: \textbf{Search space large}---fp64 $\rightarrow$ fp32 yields 14.3$\times$ (v4), but Tensor Core (tf32) is \textit{worse} than fp32.
\end{itemize}

\textbf{Conclusion:} Traditional tile/layout auto-tuning offers zero benefit for this workload class. All gains come from \textbf{fusion + precision + parallel mapping}, which are architectural decisions, not auto-tunable knobs.

\subsubsection{Advanced GPU Kernel Techniques}

We systematically evaluated five SOTA GPU optimization techniques and found four to be inapplicable or harmful (Table~\ref{tab:techniques}). These are \textbf{not} implementation failures but structural mismatches: each technique targets a throughput-bound large-matrix scenario, while our workload is latency-bound with micro-matrices.

\begin{table}[h]
\caption{Systematic evaluation of advanced GPU kernel techniques}\label{tab:techniques}
\begin{tabular}{@{}lp{3.5cm}p{2.5cm}p{4cm}@{}}
\toprule
Technique & Design Target & Applicable? & Rationale \\
\midrule
Tensor Core / tf32 & Large-matrix MAC throughput & \textbf{No}---empirically worse & v4: tf32 slower than fp32 + $\Delta X=4.7$\,mm. $15\times15$ padded to 16 wastes TC tiles. \\
TMA & $128\times128$+ async global$\leftrightarrow$shared DMA & \textbf{No}---no leverage & Working set is $\sim$KB, fully register-resident. No DMA to overlap; descriptor setup cost $>$ benefit. \\
Warp Specialization & Producer-consumer pipeline (load/compute/store) & \textbf{No}---no leverage & Only 4 active warps; no memory phase to hide. Intra-step dependency chain prevents warp pipelining. \\
Persistent Kernel & Eliminate per-launch scheduling & \textbf{Yes}---already implemented & v3/v5 are single-launch kernels. Online streaming (100\,Hz real-time) is future work. \\
Block-Diagonal Stacking & Batched small GEMM via single large matmul & \textbf{No}---anti-pattern & Stacking $B$ independent matrices into block-diagonal computes $B\times$ zero off-diagonal FLOPs, strictly worse than the grid-mapping approach (v5). \\
\botrule
\end{tabular}
\end{table}

\textbf{The only technique with marginal potential} is exploiting $F$'s block-sparse structure: $F = I + \text{few } 3\times3$ off-diagonal skew-symmetric blocks. Currently computed as a dense $15\times15$ matrix, switching to block-sparse would reduce register pressure by eliminating zero-structure storage, potentially increasing occupancy and extending the batch linear-scaling range beyond $\sim$312 trajectories. However, this is a second-order throughput optimization; the three primary axes already capture the dominant gains.

\textbf{Bottom line:} These techniques address ``the compute units are starved for arithmetic'' (throughput-bound), while our problem is ``the compute units are waiting on dependencies'' (latency-bound, SM throughput 0.27\%). The three contributions---fusion, mixed precision, and batch parallelism---have already extracted all available gains from the latency-bound sequential recursion. Further improvement requires breaking the sequential dependency itself (e.g., associative-scan parallel filtering), which we discuss as future work.

%% ============================================================
%% Section 4: Evaluation
%% ============================================================
\section{Evaluation}\label{sec4}

\subsection{Throughput Comparison}\label{subsec:throughput}

Table~\ref{tab:versions} presents the complete throughput evolution. The multiplicative nature of our three contributions is evident:

\begin{equation}
\text{Total speedup} = \underbrace{8.8\times}_{\text{fusion}} \times \underbrace{14.3\times}_{\text{fp32}} \times \underbrace{148.7\times}_{\text{batch } B=256} \approx 18{,}700\times
\label{eq:multiplicative}
\end{equation}

The measured aggregate speedup at $B=256$ is 18,210$\times$ (34.0M / 1,866), consistent with the multiplicative model after accounting for measurement variance and sub-linear batch scaling beyond SM count.

\subsection{Accuracy Preservation}\label{subsec:accuracy}

\begin{table}[h]
\caption{Positioning accuracy across precision variants ($N=166{,}667$)}\label{tab:accuracy}
\begin{tabular}{@{}lrrrr@{}}
\toprule
Version & RMSE X (mm) & RMSE Y (mm) & RMSE Z (mm) & $\rho$ (\%) \\
\midrule
v0 CPU (golden) & 1017.84 & 16.83 & 8.14 & 0.1755 \\
v3 Triton fp64 & 1017.84 & 16.84 & 8.27 & 0.1755 \\
v4 Triton fp32 & 1017.73 & 16.84 & 8.27 & 0.1757 \\
v4 Triton tf32 & 1022.55 & 17.24 & 8.58 & 0.1763 \\
\botrule
\end{tabular}
\end{table}

The fp32 fused kernel introduces a maximum deviation of $\Delta X \approx 0.11$\,mm from the fp64 golden standard---well within millimeter-level positioning requirements. The dominant X-axis error ($\sim$1\,m RMSE) stems from systematic SINS integration drift, not numerical precision, making fp32 a ``free lunch'' for this application.

\subsection{SM Utilization and Scaling}\label{subsec:sm_util}

The most compelling evidence for our approach comes from hardware utilization metrics:

\begin{itemize}
\item \textbf{Single trajectory (v4):} \texttt{sm\_\_throughput} = 0.27\%, \texttt{sm\_\_warps\_active} = 6.25\% (4/64 warps), 1/78 SMs occupied. These numbers quantify the ``GPU-hostile'' nature of the workload.
\item \textbf{Batch $B=78$ (v5):} \texttt{sm\_\_throughput} rises to 19.91\% ($\sim$74$\times$ increase), with 78 blocks occupying all SMs. This validates trajectory-dimension mapping as the correct parallelism strategy.
\item \textbf{Occupancy headroom:} 110 registers/thread $\rightarrow$ occupancy limit of 4 blocks/SM $\rightarrow$ theoretical saturation at $\sim$312 trajectories. The sub-linear scaling at $B=256$ (148.7$\times$ vs.~ideal 256$\times$) is consistent with approaching this register-pressure limit.
\end{itemize}

\subsection{Per-Step Latency Decomposition}

To contextualize the fusion gain, we decompose the per-step cost across implementations:

\begin{itemize}
\item \textbf{cuBLAS GEMM only} ($F \cdot P \cdot F^T$, $15\times15$): 17.5\,$\mu$s
\item \textbf{cuSOLVER inv only} ($3\times3$): 134\,$\mu$s
\item \textbf{v0 GPU eager} (full step, 531 kernels): $\sim$817\,$\mu$s
\item \textbf{v3 Triton mega-kernel} (full step, 1 kernel): $\sim$49\,$\mu$s
\end{itemize}

The mega-kernel's per-step cost of $\sim$49\,$\mu$s is \textbf{2.7$\times$ faster than a single cuSOLVER $3\times3$ inverse call} (134\,$\mu$s). This vividly illustrates why library-based approaches fail for this workload: the library's fixed overhead exceeds the entire fused computation.

%% ============================================================
%% Section 5: Related Work
%% ============================================================
\section{Related Work}\label{sec5}

\subsection{GPU Acceleration of Kalman Filters}

GPU acceleration of Kalman filters has been explored primarily in the context of batched estimation: large numbers of independent low-dimensional filters running concurrently~\cite{bib4}. Libraries such as cuBLAS batched GEMM, cuSOLVER batched decomposition, and MAGMA provide optimized primitives for this pattern. However, these approaches target the \textit{batched independent} scenario; single-trajectory latency-bound recursive filtering has received far less attention. Our work fills this gap by demonstrating that kernel fusion, not batched library calls, is the correct strategy for single-trajectory sequential EKF acceleration.

\subsection{Kernel Fusion for Sequential Computation}

The idea of fusing sequential computation into single GPU kernels has been successfully applied in several domains. FlashAttention~\cite{bib5} fuses attention computation to avoid materializing the full attention matrix. The Mamba selective-scan~\cite{bib6} and S5/S4 state-space model implementations use associative scans within Triton kernels for efficient sequential processing. Our work applies the same philosophy---keep state resident on-chip, eliminate intermediate global memory round-trips---to the EKF recursion, with the additional challenge that Kalman steps involve heterogeneous operations (trigonometric functions, matrix multiplies, inverses, and conditionals) that must coexist within a single kernel.

\subsection{Associative-Scan Parallel Filtering}

S\"arkk\"a and Garc\'ia-Fern\'andez~\cite{bib7} introduced temporal parallelization of Bayesian smoothers by reformulating the Kalman recursion as an associative binary operator compatible with parallel prefix-sum (Blelloch scan), reducing sequential depth from $O(N)$ to $O(\log N)$. Yaghoobi et al.~\cite{bib8} extended this to nonlinear EKF/UKF via iterative outer-loop linearization with inner-loop time-parallel filtering. This represents a qualitatively different direction from our work: while we optimize the constant factor of the $O(N)$ sequential recursion, associative-scan methods attack the asymptotic complexity. PyTorch's lack of a native \texttt{associative\_scan} primitive and the memory cost of materializing $N$ intermediate messages ($\sim$GB for $N=166{,}667$, 15-state) remain practical obstacles.

\subsection{Mixed Precision in Scientific Computing}

Mixed-precision techniques have been extensively studied for large-scale linear algebra and deep learning~\cite{bib9}. Our contribution is demonstrating that fp32 is not merely ``good enough'' but \textit{preferable} for this class of micro-matrix recursive computation: the precision loss is negligible ($\ll$ sensor noise), while the speedup (14.3$\times$ over fp64 in the fused kernel) far exceeds the fp64:fp32 ALU ratio due to compounding effects on register pressure and transcendental function latency.

%% ============================================================
%% Section 6: Conclusion
%% ============================================================
\section{Conclusion}\label{sec6}

We have presented a systematic study of GPU acceleration for a latency-bound, micro-matrix recursive workload: the SINS/dual-odometer EKF fusion positioning simulation for cased-hole robots. Our key findings are:

\begin{itemize}
\item \textbf{The problem is latency-bound, not throughput-bound.} SM throughput of 0.27\% and 531 kernel launches per step make naive GPU execution 4$\times$ slower than CPU. Traditional auto-tuning and batched library approaches are structurally mismatched.

\item \textbf{Kernel fusion is the primary lever} for latency reduction. Collapsing 531 micro-kernels into a single Triton mega-kernel with register-resident state achieves 8.8$\times$ speedup (fp64).

\item \textbf{Mixed precision (fp32) provides a ``free lunch''}: 14.3$\times$ additional speedup over the fused fp64 kernel while maintaining sub-millimeter accuracy. Tensor Cores (tf32) are counterproductive for micro-matrices.

\item \textbf{Multi-trajectory batch parallelism saturates idle SMs}: perfect linear scaling to the SM count (78$\times$), and 148.7$\times$ at $B=256$, yielding 34 million aggregate steps/s---18,210$\times$ over CPU baseline.

\item \textbf{The three contributions are multiplicative}: total speedup $\approx$ fusion $\times$ precision $\times$ batch parallelism $=$ 8.8$\times$ $\times$ 14.3$\times$ $\times$ 148.7$\times$ $\approx$ 18,700$\times$.
\end{itemize}

Our systematic negative results---Tensor Cores, TMA, warp specialization, tile/layout auto-tuning---establish that the optimization space for latency-bound micro-matrix recursive workloads is structurally distinct from that of throughput-bound large-matrix workloads. Practitioners facing similar workloads (particle filters, sensor fusion, sequential Monte Carlo) should prioritize kernel fusion, precision selection, and independent-trajectory parallelism over conventional GPU optimization recipes.

\subsection{Future Work}\label{subsec:future}

Several directions extend this work:

\begin{enumerate}
\item \textbf{Associative-scan parallel EKF}: Breaking the $O(N)$ sequential dependency via parallel prefix-sum filtering would qualitatively change the single-trajectory latency asymptotics. This requires either implementing Blelloch scan in Triton or porting to JAX for its native \texttt{associative\_scan} primitive.

\item \textbf{Block-sparse $F$ matrix exploitation}: The state transition matrix $F = I + \text{few } 3\times3$ off-diagonal blocks. Exploiting this sparsity could reduce register pressure, increase occupancy from the current 110 registers/thread, and extend the batch linear-scaling range beyond $\sim$312 trajectories.

\item \textbf{Persistent kernel with online streaming}: For real-time 100\,Hz deployment, a persistent kernel that reads sensor data from a ring buffer and maintains state across frames would eliminate per-frame launch overhead.

\item \textbf{Real sensor validation}: All experiments use simulated data. Validating the fp32 optimization on real downhole sensor recordings would strengthen the practical applicability claim.
\end{enumerate}


\backmatter

\bmhead{Acknowledgements}

[To be added.]

\section*{Declarations}

\begin{itemize}
\item Funding: [To be added.]
\item Conflict of interest: The authors declare no conflict of interest.
\item Data availability: Simulation data and code are available upon request.
\end{itemize}

\bibliography{sn-bibliography}

\end{document}